\section{Dissipative Hamiltonian Systems} \label{sec:2}
We first summarize needed concepts around the geometry of a symplectic linear vector space and then discuss Hamiltonian systems subject to dissipative forces. Finally, we introduce the Hamiltonian extension of dissipative Hamiltonian systems.

\subsection{Hamiltonian Systems} \label{sec:2.1} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$. Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
\begin{equation} \label{eq:2.1}
	\begin{aligned}
	\dot {z}(t) &= \mathbb J_{2n} \nabla_{z} H, \\
	z(0) &= z_0,
	\end{aligned}
\end{equation}
where $z \in\mathbb R^{2n}$ is the state coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$, for all $x,y\in \mathbb R^{2n}$ \cite{Marsden:2010:IMS:1965128}. By using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} method, one can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
\begin{equation} \label{eq:2.2}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix},
\end{equation}
where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$, respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
\begin{theorem} \label{theorem:2.1}
\cite{Marsden:2010:IMS:1965128} Consider the flow $\phi_t:\mathbb R\times \mathbb R^{2n} \to \mathbb R^{2n}$ of the Hamiltonian system (\ref{eq:2.1}). Then $H\circ \phi_t = H$.
\end{theorem}

In many physical problems, $H$ represents the system energy and is bounded from below. Here, to avoid difficulties with well-posedness of Hamiltonian systems,  we often assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix. This assumption leads to a linear system of evolution (\ref{eq:2.1}). We emphasise that the Hamiltonian extension in Section \ref{sec:2.2} and the Reduced Dissipative Hamiltonian method in Section \ref{sec:3.3} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$. 

Under general coordinate transformations, the equations of evolution may not take the form in (\ref{eq:2.1}). Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces. A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
\begin{equation}
	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
\end{equation}
In matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
\begin{equation}
	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
\end{equation}
where the superscript $T$ represents the transpose operator. As the symplectic 2-form is preserved under a symplectic transformation, the form of the equations of evolution remains invariant through a symplectic coordinate transformation \cite{Marsden:2010:IMS:1965128}. The \emph{symplectic inverse} of $A$, is a pseudo-inverse given by
\begin{equation}
	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
\end{equation}
It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix. Furthermore, one easily checks that $AA^+$ is idempotent, i.e. it is a projection operator onto the column span of $A$. 

It is known that symplectic matrices are usually ill-conditioned \cite{Karow:2006cf}. Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded. A basis which is both symplectic and orthogonal is called a \emph{ortho-symplectic basis}. We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.

It is natural to expect the numerical integrator that solves (\ref{eq:2.1}) also satisfy the conservation law in Theorem \ref{theorem:2.1}. However, common numerical integrators, e.g. the Runge-Kutta method, do not generally preserve the Hamiltonian. Symplectic numerical integrators are a class of numerical integrators for Hamiltonian systems that preserves the symplectic structure and ensure stability during long-time integration. The Str\"omer-Verlet time-stepping scheme
\begin{equation} \label{eq:2.3}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}),
\end{aligned}
\end{equation}
where $p$ and $q$ are the canonical coordinates $z = (q^T,p^T)^T$, is an example of such numerical integrators. More information on the construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.

\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions} \label{sec:2.2}

Many systems in engineering and science appear as a perturbation of a Hamiltonian system, where the perturbation can be regarded as dissipation. In these systems, the energy tends to decrease over time, and thus, the conservation law in Theorem \ref{theorem:2.1} does not hold. Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system. 

To account for dissipation in a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
\begin{equation} \label{eq:2.4}
	\begin{aligned}
		& \dot {z} = \mathbb J_{2n} K^T f(t), \\
		& z(0) = z_0,
	\end{aligned}
\end{equation}
where $f$ is the solution to the Volterra integral equation \cite{corduneanu1991integral}
\begin{equation} \label{eq:2.4.1}
	f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K z.
\end{equation}
Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}. Note that the integral term in (\ref{eq:2.4.1}) accounts to the accumulation of the dissipation, whereas $\chi(s) = 0$ implies (\ref{eq:2.4}) is equivalent to (\ref{eq:2.1}). Furthermore, under suitable assumptions on $K$, both (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.

\begin{example} \label{example:2.1}
Consider the dynamics of the damped harmonic oscillator
\begin{equation} \label{eq:2.5}
	\ddot q + r \dot q + k q = 0
\end{equation}
where $k$ is the Hooke's constant and $r$ is the spring's damping factor. Note that without a damping term, (\ref{eq:2.5}) is a Hamiltonian system. The TDD formulation for the damped harmonic oscillator takes the form
\begin{equation} \label{eq:2.6}
	\dot q(t) = f(t), \quad \dot p(t) = - k q(t), \quad f(t) + \int_0^t r f(s) \ ds = p(t).
\end{equation}
Here $(q,p)$ are the canonical coordinates and the susceptibility is the constant function $r$.
\end{example}

It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on the linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense. In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, and $x$ is the solution to
\begin{equation} \label{eq:2.7}
	\dot x = \mathcal J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
\end{equation}
Here $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function and $\mathcal J_{2n}$ is the symplectic operator defined on $\mathbb R^{2n}\times \mathcal H^{2n}$ respectively.

\begin{theorem}
Suppose that $K$ is full rank and $\chi(t)$ is symmetric. Then there is a quadratic extension to (\ref{eq:2.4}) of the form (\ref{eq:2.7}), if
\begin{equation} \label{eq:2.8}
	\text{Im}(\xi\hat{\chi}(\xi)) \geq 0, \quad \forall \xi = \omega + i\eta, \ \eta \geq 0,
\end{equation}
where $\hat{\chi}$ is the Fourier-Laplace transform of $\chi$
\begin{equation} \label{eq:2.9}
	\hat{\chi}(\xi) = \int_0^\infty e^{i\xi t} \chi(t)\ dt.
\end{equation}
\end{theorem}
\emph{Proof.} Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite. We refer the reader to \cite{Figotin:2006jy} for the proof of the general case. Consider the Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x), \\
		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), 
\end{align}
\end{subequations}
together with the initial condition
\begin{equation} \label{eq:2.10.1}
	z(0) = z_0,\quad \phi(0,\cdot) = 0, \quad \theta(0,\cdot) = 0.
\end{equation}
Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
\begin{equation} \label{eq:2.11}
	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
\end{equation}
To show that the Hamiltonian system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}). Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are equations for a vibrating string, and can be solved analytically
\begin{equation} \label{eq:2.12}
	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
\end{equation}
Then, we recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) takes the quadratic from
\begin{equation} \label{eq:2.13}
	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
\end{equation}
where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$. 

Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are called the \emph{hidden strings}. The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings making the extended system conservative. The Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900} which is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.

Note that the time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}). In general, the history of $f(t)$ must be stored and may cause storage limitation in long-time integration. However, we are interested solely in finding $z(t)$ which depends on $f$ at time $t$, and $\phi(t,0)$, i.e. the integral of the history of $f$. So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$. For example for the trapezoidal rule, we recover the recursive relation
\begin{equation} \label{eq:2.14}
	\int_{0}^{t_n} f(s) \ ds \approx \frac{\Delta t}{2} f(t_n) + \frac{\Delta t}{2} f(t_{n-1}) + \int_{0}^{t_{n-1}} f(s) \ ds,
\end{equation}
where $\Delta t$ is the time step. The recursive relation in (\ref{eq:2.14}) suggests that storing the value of the integral term together with the state of $f$ in the previous time step suffices to evaluate the integral for the new time step. For other interpolation based quadrature rules, we can construct similar recursive rules of the form
\begin{equation}
	\int_{0}^{t_n} f(s) \ ds \approx \sum_{i=0}^{k} \omega_i f(t_{n-i})  + \int_{0}^{t_{n-k}} f(s) \ ds
\end{equation}
for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), only requires storage of $k$ evaluations of $f$.
